---
output: github_document
bibliography: inst/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/README-",
  out.width = "100%"
)
```

# Point Process Principal Component Analysis

<!-- badges: start -->
<!-- badges: end -->

> :warning: The `{pppca}` package is under final development stage (documentation, unit testing, etc.). You can already use it as the code is not expected to change much before its official release. :warning:

The `{pppca}` R package provides an implementation of the Principal Component Analysis (PCA) framework for point process data.

See Picard et al (2024) for more details about the method.

## Authorship and license

This package is developed by:

- Franck Picard ([link](https://franckpicard.github.io/)|[ORCID](https://orcid.org/0000-0001-8084-5481))
- Ghislain Durif ([link](https://gdurif.perso.math.cnrs.fr/)|[ORCID](https://orcid.org/0000-0003-2567-1401))

The `{pppca}` package is available under the GPL>=3 license (except for the included `earthquake` dataset, see `?earthquake` or the [`LICENSE`](./LICENSE) file).

## Installation

You can install the development version of `{pppca}` from [GitHub](https://github.com/) with:
```R
# install.packages("remotes")
remotes::install_github("franckpicard/pppca")
```

## Usage

See the tutorial [below](#below) made with the vignette ["PCA for Point Processes with Earthquake Data"](vignettes/01-introduction-pppca.Rmd):
```R
vignette("01-introduction-pppca", package = "pppca")
```

## Note for developpers

The `{pppca}` package is developed using the [`{fusen}`](https://thinkr-open.github.io/fusen/index.html) framework to generate R packages. Check the [`dev`](./dev) folder in the git repository for the sources and recipe.

See the [`TODO.md`](./TODO.md) file in the git repository for information about ongoing development work.

# Tutorial

```{r vignette-intro}
#| child: vignettes/01-introduction-pppca.Rmd
```
