---
title: "Development actions history"
output: html_document
editor_options: 
  chunk_output_type: console
---

All commands that you use to use when developing packages...

# First time just after creating the project

- Fill the following chunk to create the DESCRIPTION of your package

```{r dev-description, eval=FALSE}
# Describe your package
fusen::fill_description(
  pkg = here::here(),
  fields = list(
        Title = stringr::str_to_title(
            "Point Process Principal Component Analysis (PPPCA)"),
        Description = 
            "Implementation of Principal Component Analysis (PCA) 
            for point process data.",
        `Authors@R` = c(
            person(
                "Franck", "Picard", 
                email = "franck.picard@ens-lyon.fr", 
                role = c("aut"), 
                comment = c(ORCID = "0000-0001-8084-5481")),
            person(
                "Ghislain", "Durif", email = "gd.dev@libertymail.net", 
                role = c("aut", "cre"), 
                comment = c(ORCID = "0000-0003-2567-1401"))
        ),
        URL = "https://github.com/franckpicard/pppca",
        Version = "0.0.1",
        "Date/Publication" = format(Sys.time(), "%Y/%m/%d")
    )
)
# Define License with use_*_license()
usethis::use_mit_license("John Doe")
# Roxygen markdown
usethis::use_roxygen_md()
```

# All-in-one function to deploy publicly on GitHub

Either use this GitHub all-in-one function or choose the steps one-by-one in the following sections.
See `vignette("share-on-a-github-website", package = "fusen")` for more information.

```{r, eval=FALSE}
# _All-in-one share of your package and its documentation on GitHub
fusen::init_share_on_github()
```


# Start using git

```{r dev-git, eval=FALSE}
# Deal with classical files to ignore
usethis::git_vaccinate()
```

# Set extra sources of documentation

```{r dev-extra-doc, eval=FALSE}
# README
usethis::use_readme_rmd()
```

**From now, you will need to "inflate" your package at least once to be able to use the following commands. Let's go to your flat template, and come back here later if/when needed.**


# Package development tools

## Use once

```{r dev-extra-req, eval=FALSE}
# Pipe
usethis::use_pipe()

# package-level documentation
usethis::use_package_doc()
```

## Inflate package

```{r dev-inflate}
fusen::inflate_all()
```

## Use everytime needed

```{r dev-doc, eval=FALSE}
# Simulate package installation
pkgload::load_all()

# Generate documentation and deal with dependencies
attachment::att_amend_desc()
```

```{r dev-check}
# Test the package
devtools::test()
```

```{r dev-check}
# Check the package documentation
devtools::check_man()
```

```{r dev-check}
# Check the package
devtools::check()
```

# Share the package

```{r, eval=FALSE}
# set and try pkgdown documentation website locally
usethis::use_pkgdown()
pkgdown::build_site()

# build the tar.gz with vignettes to share with others
devtools::build(vignettes = TRUE)

# Share your package and its documentation on GitHub
# usethis::create_github_token()
# gitcreds::gitcreds_set()
fusen::init_share_on_github()
```

