---
title: "PCA for Point Processes with Earthquake Data"
author: Franck
date: May 24
output:
 html_document:
   toc: true
   toc_depth: 3
   number_sections: true
   highlight: tango
   latex_engine: xelatex 
   fig_caption: true
   
---

```{r, message=FALSE,fig.show="show",eval=TRUE,warning=FALSE,echo=FALSE}
library(ggplot2)
library(ggpubr)
library(purrr)
library(wesanderson)
library(parallel)
my_theme <- function(){
  theme(   
    axis.line        = element_line(colour = "black"),
    strip.background = element_blank(),
    panel.border     = element_rect(colour = "black", fill = NA),
    panel.background = element_rect(fill=NA),
    strip.text.x     = element_text(size = 10),
    strip.text.y     = element_text(size = 10),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    legend.position  = "top",
    legend.title     = element_blank(),
    legend.text      = element_text(size=6),  
    aspect.ratio     = 1,
    axis.text.x      = element_text(size=5),
    axis.text.y      = element_text(size=5)
  )
}
source("/home/picard/projets/pppca/pppca_fun.R")
load(file="/home/picard/Dropbox/PCAforPointProcesses/numerical_study/vignette/earthquake_data.RData")
PP_num = lapply(PP, as.numeric)
Jmax = 50
res  = pppca(PP_num,Jmax)

```

# Data

The data were obtained from the Kandilli Observatory and Earthquakes Research Institute at Bo\u{g}azi\c{c}i University\footnote{\url{http://www.koeri.boun.edu.tr/sismo/2/earthquake-catalog/}}. The dataset comprises earthquake occurrences in Turkey and neighboring regions of Greece, recorded between January 2013 and January 2023, spanning 1181 cities. In recent years, the Gulf of G\"{o}kova in Southwest Turkey has witnessed two significant seismic events: the Bodrum earthquake on July 20, 2017 (magnitude 6.6, also felt on the Greek island of Kos), and the Aegean Sea earthquake on November~1, 2020, with a moment magnitude of 7.0 (the highest magnitude observed during the period).  In the following, we illustrate how our framework can be employed to investigate the fine-scale dynamics of earthquake occurrences in the region, focusing on the 195 cities that experienced more than two earthquakes over the specified period.	

# Model 

The earthquakes occurrences are modeled with $n$ independent and identically distributed (i.i.d.) temporal point processes $(N_1, N_2, \ldots, N_n)$,  observed on the time interval $[0,1]$, such that:
$$
N^i = (T^i_1, T^i_2,...), \quad T^i_1< T^i_2 < ...
$$

```{r, message=FALSE,fig.show="show",eval=TRUE,warning=FALSE,echo=FALSE}
mat = lapply(1:length(PP), FUN=function(i){data.frame(y=i,x=PP[[i]])})
mat = Reduce("rbind",mat)
gg1 = ggplot(mat, aes(x=x,y=y)) + geom_point(size=0.001)+ my_theme() + 
  xlab("Date") + 
  ylab("City") + 
  theme(axis.text.y=element_blank(),axis.ticks.y=element_blank()) + 
  geom_vline(xintercept=as.POSIXct("2017.07.16",format = "%Y.%m.%d"),color = "gray") +
  geom_vline(xintercept=as.POSIXct("2020.11.01",format = "%Y.%m.%d"),color = "gray") 

# for the first 50 cities in the dataset
pmf = get_pmf(PP)
mat = lapply(1:50, FUN=function(i){
  data.frame(city=names(pmf)[i],date=as.POSIXct(res$grid[-1]),pmf = pmf[[i]])
})
mat = Reduce("rbind",mat)
gg2 = ggplot(mat, aes(x=date,y=pmf, group=city)) + geom_step()+ my_theme() + 
  xlab("Date") + 
  ylab("Cumulative Mass Function") + 
  theme(axis.text.y=element_blank(),axis.ticks.y=element_blank()) + 
  geom_vline(xintercept=as.POSIXct("2017.07.16",format = "%Y.%m.%d"),color = "gray") +
  geom_vline(xintercept=as.POSIXct("2020.11.01",format = "%Y.%m.%d"),color = "gray") 

ggarrange( gg1+theme(plot.margin = unit(c(0,0,0,0), 'mm')),
           gg2+theme(plot.margin = unit(c(0,0,0,0), 'mm')),
           ncol = 2, nrow = 1)
```

Our method is based on the associated \textit{cumulative mass functions} of the processes (called counting processes), such that 
$$
\Pi_i([0,t]) = \sum_{T \in N_i} 1_{ \left\{ T<t \right\}}.
$$

# PCA for Point Processes
	
As in classical PCA, we first inspect the percentage of variance carried by each axis, the first axis carrying  $\widehat{\lambda}_1/\sum_j \widehat{\lambda}_j = 89\%$ of the variability. This implies that the first source of variance in the data is associated with the deviation of cities in the accumulation of earthquakes over time compared to the average temporal pattern of earthquakes in the region. 

```{r, message=FALSE,fig.show="show",eval=TRUE,warning=FALSE,echo=FALSE}
dd  = data.frame(percentvar = res$percentvar[1:Jmax], j = 1:Jmax)
gg1 = ggplot(dd, aes(x=j, y=percentvar)) + geom_line() + geom_point(size=0.5)+ coord_trans(y="log")+ my_theme() + xlab("Nb of eigen elements") + ylab("%variance (log)")

dd  = data.frame(axis1 = res$scores[,1], nb_occ = sapply(PP, length))
gg2 = ggplot(dd,aes(x=nb_occ,y=axis1)) + geom_point(size=0.5) + my_theme() + 
  geom_abline(slope = lm(dd$axis1~dd$nb_occ)$coefficients[2], intercept = lm(dd$axis1~dd$nb_occ)$coefficients[1]) +
  xlab("Number of Occurrences") + 
  ylab(expression(xi[1])) 

ggarrange( gg1+theme(plot.margin = unit(c(0,0,0,0), 'mm')),
           gg2+theme(plot.margin = unit(c(0,0,0,0), 'mm')),
           ncol = 2, nrow = 1)
```

Also, scores on Axis 1 (\((\widehat\xi_{i1})_i\)) directly correspond to the total number of earthquakes in each city over the specified period. We identify Akdeniz as an outlier with an unusually high number of earthquakes compared to the regional average. This city will be excluded from subsequent analyses that investigate the finer-scale dynamics of earthquake activity in the area.

```{r, message=FALSE,fig.show="show",eval=TRUE,warning=FALSE,echo=FALSE}
dd = data.frame(x=res$grid[-1],y=res$eigenfun[,1])
ggplot(dd,aes(x=x,y=y)) + 
  geom_vline(xintercept=as.POSIXct("2017.07.16",format = "%Y.%m.%d"),color = "gray") +
  geom_vline(xintercept=as.POSIXct("2020.11.01",format = "%Y.%m.%d"),color = "gray") +
  geom_line() + geom_hline(yintercept = 0,linetype=2) + my_theme() + 
  xlab("Date") +
  ylab(expression(eta[1]))
```

# Interpretation of scores

In order to interpret the subsequent axes, we recall that our method summarizes the dynamics of earthquake occurrences through simple functions, as expressed in 
$$
\Pi_i([0,t]) \approx W([0,t])+ \sum_{j=1}^J\sqrt{\lambda_j} \widehat{\xi}_{i,j} \widehat{\eta}_{j}(t), \qquad t\geq 0
$$
Consequently, we focus on the variations of the estimated eigenfunctions $(\widehat{\eta}_j)_j$, and represent the positions of cities according to their scores $(\widehat{\xi}_j)_j$ on successive axes as in any PCA analysis. These representations allow us to identify cities (like Gokova Korfezi, Onika Adalar) that have typical accumulations of earthquakes over the period.

```{r, message=FALSE,fig.show="show",eval=TRUE,warning=FALSE,echo=FALSE}
j = 2
d = data.frame(x = res$scores[,j],y = res$scores[,j+1])
gg1 = ggplot(d,aes(x=x,y=y,label=names(PP))) + geom_point(size=0.5) + my_theme() + geom_hline(yintercept = 0,linetype=2) + geom_vline(xintercept = 0,linetype=2) + 
geom_text(aes(label=ifelse(x>2,names(PP),'')),nudge_y = 0.2, size=2)+
geom_text(aes(label=ifelse(x<(-2),names(PP),'')),nudge_y = 0.2, size=2)+
geom_text(aes(label=ifelse(y>1,names(PP),'')),nudge_y = 0.2, size=2)+
geom_text(aes(label=ifelse(y<(-2),names(PP),'')),nudge_y = 0.2, size=2)+
xlab(expression(xi[2])) +
ylab(expression(xi[3])) +
xlim(-10, 10)

j = 4
d = data.frame(x = res$scores[,j],y = res$scores[,j+1])
gg2 = ggplot(d,aes(x=x,y=y,label=names(PP))) + geom_point(size=0.5) + my_theme() + geom_hline(yintercept = 0,linetype=2) + geom_vline(xintercept = 0,linetype=2) + 
geom_text(aes(label=ifelse(x>4,names(PP),'')),nudge_y = 0.2, size=2)+
geom_text(aes(label=ifelse(x<(-2),names(PP),'')),nudge_y = 0.2, size=2)+
geom_text(aes(label=ifelse(y>1,names(PP),'')),nudge_y = 0.2, size=2)+
geom_text(aes(label=ifelse(y<(-2),names(PP),'')),nudge_y = 0.2, size=2)+
xlab(expression(xi[4])) +
ylab(expression(xi[5])) +
xlim(-10, 10)

ggarrange( gg1+theme(plot.margin = unit(c(0,0,0,0), 'mm')),
           gg2+theme(plot.margin = unit(c(0,0,0,0), 'mm')),
           ncol = 2, nrow = 1)


```

# Interpretation of eigenfunctions

Eigenfunctions can be interpreted as the basic components that drive the dynamics of earthquake occurrences over time. Interestingly, the second axis, \(\widehat\eta_2\) reveals a distinct change in seismic activity between the two main earthquakes (2017 and 2020). Cities exhibiting positive scores on this axis indicate a global increase in earthquake rates between July 2017 and November 2020, followed by a decrease below the average regional rate after November 2020. Conversely, cities with negative scores on \(\widehat{\eta}_3\) show an accumulation of earthquakes before July 2017. Overall, our method provides a highly accurate description of the variability in earthquake occurrences among different cities. It offers a means to represent and position cities relative to each other based on their earthquake dynamics variability. Moreover, our estimation framework, relying on occurrence data without smoothing, captures sharp and fine-scale variations in this dynamics, as shown by the distinct peaks in $\widehat{\eta}_4$ and $\widehat{\eta}_5$ 

```{r, message=FALSE,fig.show="show",eval=TRUE,warning=FALSE,echo=FALSE}
dd  = data.frame(x=res$grid[-1],y=res$eigenfun[,2])
gg1 = ggplot(dd,aes(x=x,y=y)) + 
  geom_vline(xintercept=as.POSIXct("2017.07.16",format = "%Y.%m.%d"),color = "gray") +
  geom_vline(xintercept=as.POSIXct("2020.11.01",format = "%Y.%m.%d"),color = "gray") +
  geom_line() + geom_hline(yintercept = 0,linetype=2) + my_theme() + 
  xlab("Date") +
  ylab(expression(eta[2])) +
  theme(axis.text.y= element_text(angle = -90, vjust = 0.5, hjust=1))

dd  = data.frame(x=res$grid[-1],y=res$eigenfun[,3])
gg2 = ggplot(dd,aes(x=x,y=y)) + 
  geom_vline(xintercept=as.POSIXct("2017.07.16",format = "%Y.%m.%d"),color = "gray") +
  geom_vline(xintercept=as.POSIXct("2020.11.01",format = "%Y.%m.%d"),color = "gray") +
  geom_line() + geom_hline(yintercept = 0,linetype=2) + my_theme() + 
  xlab("Date") +
  ylab(expression(eta[3]))+
  theme(axis.text.y= element_text(angle = -90, vjust = 0.5, hjust=1))

dd  = data.frame(x=res$grid[-1],y=res$eigenfun[,4])
gg3 = ggplot(dd,aes(x=x,y=y)) + 
  geom_vline(xintercept=as.POSIXct("2017.07.16",format = "%Y.%m.%d"),color = "gray") +
  geom_vline(xintercept=as.POSIXct("2020.11.01",format = "%Y.%m.%d"),color = "gray") +
  geom_line() + geom_hline(yintercept = 0,linetype=2) + my_theme() + 
  xlab("Date") +
  ylab(expression(eta[4]))+
  theme(axis.text.y= element_text(angle = -90, vjust = 0.5, hjust=1))

dd  = data.frame(x=res$grid[-1],y=res$eigenfun[,5])
gg4 = ggplot(dd,aes(x=x,y=y)) + 
  geom_vline(xintercept=as.POSIXct("2017.07.16",format = "%Y.%m.%d"),color = "gray") +
  geom_vline(xintercept=as.POSIXct("2020.11.01",format = "%Y.%m.%d"),color = "gray") +
  geom_line() + geom_hline(yintercept = 0,linetype=2) + my_theme() + 
  xlab("Date") +
  ylab(expression(eta[5]))+
  theme(axis.text.y= element_text(angle = -90, vjust = 0.5, hjust=1))

ggarrange( gg1+theme(plot.margin = unit(c(0,0,0,0), 'mm')),
                gg2+theme(plot.margin = unit(c(0,0,0,0), 'mm')),
                gg3+theme(plot.margin = unit(c(0,0,0,0), 'mm')),
                gg4+theme(plot.margin = unit(c(0,0,0,0), 'mm')),
                ncol = 2, nrow = 2)
```

